---
title: "relation"
output: html_document
date: "2023-10-23"
---

```{r}
cvdnutr <- sig_biom_final %>% 
            left_join(sig_nutr_final %>% select(-group), by = "ABSPID") 
```



Lets define high risk cvd as satisfying at least two of the CVD criteria:
```{r}
cvdnutr$cvd_risk <- ifelse(
  (
      ifelse(!is.na(cvdnutr$Obesity) & cvdnutr$Obesity == "Obese", 1, 0) +
      # ifelse(!is.na(cvdnutr$Hypertension) & cvdnutr$Hypertension == "Hypertensive", 1, 0) +
      ifelse(!is.na(cvdnutr$Diabetes) & cvdnutr$Diabetes == "Diabetic", 1, 0) +
      ifelse(!is.na(cvdnutr$ApoB) & cvdnutr$ApoB == "Abnormal ApoB", 1, 0) +
      ifelse(!is.na(cvdnutr$Dyslipidemia) & cvdnutr$Dyslipidemia== "Dyslipidemia", 1, 0)) >= 2, "high", "low")
      #  + ifelse(!is.na(cvdnutr$SMKDAILY) & cvdnutr$SMKDAILY %in% c("1. Current daily smoker", "2. Past daily smoker (still smokes)"), 1, 0)
```

```{r}
table(cvdnutr$cvd_risk, cvdnutr$group)
```




```{r}
# Define risk categories in dietary variables (Nutrition students) 

cvdnutrdf = cvdnutr %>% mutate(
  Vegetables_Legumes_beans = case_when(
    VEGLEGN_mean < 5.5 ~ '< 5.5',
    VEGLEGN_mean > 5.5 ~ '≥ 5.5',TRUE ~ NA_character_),
  Fruit = case_when(
    FRUITN_mean < 2 ~ '< 2 ',
    FRUITN_mean >= 2 ~ '≥ 2 ',TRUE ~ NA_character_),
  Grains_cereals = case_when(
    GRAINSN_mean < 6 ~ '< 6 ',
    GRAINSN_mean >= 6 ~ '≥ 6 ',TRUE ~ NA_character_),
  Meat_Poultry = case_when(
    MEATN_mean < 2.5 ~ '< 2.5 ',
    MEATN_mean >= 2.5 ~ '≥ 2.5 ',TRUE ~ NA_character_),
  Dairy = case_when(
    DAIRYN_mean < 2.5 ~ '< 2.5 ',
    DAIRYN_mean >= 2.5 ~ '≥ 2.5 ',TRUE ~ NA_character_),
  Unsaturated_Spread_Oils = case_when(
    UNSATN_mean < 4 ~ '< 4 ',
    UNSATN_mean >= 4 ~ '≥ 4 ',TRUE ~ NA_character_),
  FreeSugar = case_when(
    UNSATN_mean < 10 ~ '< 10 %',
    UNSATN_mean >= 10 ~ '≥ 10 %',TRUE ~ NA_character_),
  TotalFat = case_when(
    FATPER_mean < 30 ~ '< 30 %',
    FATPER_mean >= 30 ~ '≥ 30 %',TRUE ~ NA_character_),
  Carbohydrates = case_when(
    CHOPER_mean <= 40 ~ '≤ 40 %',
    CHOPER_mean > 40 & CHOPER_mean < 70 ~ '40 -70 %',
    CHOPER_mean >= 70 ~ '≥ 70 %',TRUE ~ NA_character_),
  Protein = case_when(
    PROPER_mean <= 15 ~ '< 15 %',
    CHOPER_mean > 15 & CHOPER_mean < 25 ~ '15 -25 %',
    CHOPER_mean >= 25 ~ '≥ 25 %',TRUE ~ NA_character_),
  SaturatedTransFat = case_when(
    SATPER_mean < 11 ~ '< 11 %',
    SATPER_mean >= 11 ~ '≥ 11 %',TRUE ~ NA_character_),
  DietaryFibre = case_when(
    FIBRET_mean < 30 ~ '< 30 g',
    FIBRET_mean >= 30 ~ '≥ 30 g',TRUE ~ NA_character_),
  cvd_risk_stat = ifelse(cvdnutr$cvd_risk == 'high', 1, 0)) %>% 
  select(c(colnames(sig_biom_final), Vegetables_Legumes_beans, Fruit, Grains_cereals, Meat_Poultry, Dairy, Unsaturated_Spread_Oils, FreeSugar, TotalFat, Carbohydrates, Protein, SaturatedTransFat, DietaryFibre, cvd_risk, cvd_risk_stat))

```


Build preliminary regression model (in our presentation)
```{r}
# Split the data into training and test sets (70% training, 30% test)
splitIndex <- createDataPartition(cvdnutr$cvd_risk, p = .7, list = FALSE)
train_data <- cvdnutr[splitIndex, ]
test_data <- cvdnutr[-splitIndex, ]
train_data$cvd_risk <- ifelse(train_data$cvd_risk == "high", 1, 0)
test_data$cvd_risk <- ifelse(test_data$cvd_risk == "high", 1, 0)

# Get nutrition variable names
nutr_vars <- setdiff(colnames(sig_nutr_final), c("ABSPID", "group"))

# Fit the model
model_diet <- glm(cvd_risk ~ .,
                   data = train_data %>% select(cvd_risk, all_of(nutr_vars), group),
                   family = "binomial")


# Extract coefficients from the model
coef_summary <- summary(model_diet)$coefficients

# Predict probabilities
probabilities <- predict(model_diet, type = "response", newdata = test_data)

# Convert to binary prediction based on a threshold (0.5 is common)
predicted_values <- ifelse(probabilities > 0.5, 1, 0)

# ROC curve
roc_obj <- roc(test_data$cvd_risk, probabilities)
auc(roc_obj)
plot(roc_obj, main="ROC Curve")


# Plot residuals
plot(residuals(model_diet), main="Residuals vs Fitted", xlab="Fitted values", ylab="Residuals")
abline(h = 0, col = "red")

# Generate confusion matrix
cm <- confusionMatrix(as.factor(predicted_values), as.factor(test_data$cvd_risk))

# Print the confusion matrix and related statistics
print(cm)
```


Odds ratio plot we agreed to add after presentation (Gao originally responsible for implementing)

```{r}
# Indigenous
explanatory_food <- c( "Vegetables_Legumes_beans",'Fruit','Grains_cereals','Meat_Poultry','Dairy','Unsaturated_Spread_Oils')
explanatory_nutr <- c( 'FreeSugar','TotalFat','Carbohydrates','Protein','SaturatedTransFat','DietaryFibre')
dependent <- "cvd_risk_stat"

#Indigenous cohort

p1 = cvdnutrdf %>% filter(group == 'Indigenous') %>% or_plot(dependent, explanatory_food,table_text_size = 2.6, dependent_label = "Indigenous cohort", suffix = "")

p2 = cvdnutrdf %>% filter(group == 'Indigenous') %>% or_plot(dependent, explanatory_nutr,table_text_size = 2.6, dependent_label = "Indigenous cohort", suffix = "")


# Non-Indigenous)
p3 = cvdnutrdf %>% filter(group == 'NonIndigenous') %>% or_plot(dependent, explanatory_food,table_text_size = 2.6, dependent_label = "Non-Indigenous cohort", suffix = "")
p4 = cvdnutrdf %>% filter(group == 'NonIndigenous') %>% or_plot(dependent, explanatory_nutr,table_text_size = 2.6, dependent_label = "Non-Indigenous cohort", suffix = "")

```

# Odds Plot
```{r}
p51 = grid.arrange(p1, p2, ncol = 1, top = textGrob('', gp = gpar(fontsize = 20)))
p6 = grid.arrange(p3, p4, ncol = 1, top = textGrob('', gp = gpar(fontsize = 20)))
p51
p6
```
Results show that (p < 0.05):
Ind cohort:
Meat/Poultry (>=2.5, OR = 0.6)
Unsaturated spreads/oils (>= 4, OR = 0.59)
Protein (>=25%, OR = 0.49)


NonInd cohot:
Diet with high intake of Protein (>= 25%)corresponds to 1.34 on OR plot (p < 0.05)
Diet with hight intake of Dietary fibre (>= 30g) corresponds to 0.74 on OR plot (p < 0.05)






Build regression model with these vars / some sort of classification model? interaction effect? feedback greatly appreciated. 

(to be continued...)

```{r}
model_interaction <- glm(cvd_risk_stat ~ (Vegetables_Legumes_beans + Fruit + Grains_cereals + Meat_Poultry + Dairy + Unsaturated_Spread_Oils + FreeSugar + TotalFat + Carbohydrates + Protein + SaturatedTransFat + DietaryFibre) * group, data = cvdnutrdf, family = binomial, na.action = na.exclude)
summary(model_interaction)
```

```{r}
# Use stepwise for a 'quick' best model:
stepwise_model <- MASS::stepAIC(model_interaction, direction = "both", trace = FALSE)
summary(stepwise_model)
```





