---
title: "Stat3888 Assignment"
author: 'SID: 500505887'
date: "Date: 2023-09-01"
output: 
  html_document:
    code_folding: hide
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initialisation

## Libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(here)      # directory referencing
library(readxl)    # reading Excel files
library(janitor)   # data cleaning 
library(stringr)   # string manimpuation
library(tidyr) 
library(dplyr)
library(naniar)
library(ggplot2)
library(knitr)
library(kableExtra)
```


## Functions

Label variables as continuous and categorical
```{r}
get_type <- function(dict) 
{
  #dict <- dict_nutr
  var_names <- unique(dict$variable_name)
  var_type  <- c()
  for (j in 1:length(var_names))
  {
    # Get all dictionary lines corresponding to a varible
    # (a block of lines)
    dict_block <- dict %>% 
      filter(variable_name==var_names[j])
    
    # Take all of the characters in a block, paste them 
    # together and make all characters lower case
    block_string <- dict_block %>%
      select(-variable_name) %>%
      as.matrix() %>%
      as.vector() %>%
      paste(collapse="") %>%
      tolower()
    
    # Assign variable if we can find the word "continuous"
    # in the block otherwise assume that it is "categorical"
    var_type[j] <- block_string %>% 
      str_detect("continuous") %>% 
      ifelse("continuous","categorical") 
  }
  return(var_type)
}
```


Split the description variable in the data dictionary into value and meaning columns (* Modified from given)
```{r}
get_my_special_value_meanings <- function(dict) {
  special <- tibble(variable_name = c(),
                    value = c(),
                    meaning = c())
  
  for (i in 1:nrow(dict)) {
    variable <- dict$variable_name[i]
    description <- tolower(dict$description[i])  # Convert description to lowercase
    
    # Check if description contains "continuous"
    if (grepl("continuous", description)) {
      next  # Skip entries with "continuous" in description
    }
    
    # Check if description starts with a number followed by a period and space
    if (grepl("^\\d+\\. ", description)) {
      # Extract value and meaning using pattern matching
      values <- str_match(description, "^(\\d+\\. )(.*)")
      value <- values[2]
      meaning <- values[3]
      
      result <- tibble(variable_name = variable,
                       value = value,
                       meaning = meaning)
      
      special <- bind_rows(special, result)
    } else {
      # If description doesn't follow the pattern, treat the whole description as meaning
      result <- tibble(variable_name = variable,
                       value = NA,
                       meaning = description)
      
      special <- bind_rows(special, result)
    }
  }
  
  return(special)
}
```

Split the description variable in the data dictionary into value and meaning columns (* Original)
```{r}
get_special_value_meanings <- function(dict)
{
  var_names <- unique(dict$variable_name)
  special   <- tibble(variable_name=c(),
                      value=c(),
                      meaning=c())
  
  for (j in 1:length(var_names)) 
  {
    # Get a block of values from the dictionary
    block <-  dict %>%
      filter(variable_name==var_names[j])
  
    if (nrow(block)>1) {
      # Split  the descrition into value/meaning pairs
      special_block <- block[-1,-c(3:ncol(block))] %>%
        dplyr::filter(!grepl("continuous",tolower(description))) %>%
        separate(col=2, 
                 sep="[.]",
                 into=c("value","meaning")) %>%
        mutate(value=as.numeric(value),
               meaning=tolower(str_trim(meaning))) 
      
      # append these to a block of special value/meaning pairs
      special <- bind_rows(special, special_block)
    }
  }
  return(special)
}
```


Label all categorical variables and convert them to factors in R.
```{r}
categorical_to_factor <- function(types, proc) 
{
  var_names <- colnames(proc)  
  for (i in 1:length(var_names)) 
  {
    # Extract the inferred variable type from the types tibble
    var_type <- types %>% 
      filter(variable_name==var_names[i]) %>%
      select(variable_type) %>%
      as.character()
    
    # If the type is categorical turn the variable into a factor
    if (var_type=="categorical") {
      proc[[var_names[i]]] <- as.factor(proc[[var_names[i]]])
    }
  }
  return(proc)
}
```


Convert raw data to technically correct data
```{r}
miss_defs <- c("Did not proceed to Biomedical component",
                     "Measurement not taken - equipment faulty",
                     "Measurement not taken - other reason",
                     "Measurement not taken - refusal",
                     "None of these apply",
                     "Not applicable",
                     "Not collected",
                     "Not currently on a diet",
                     "Not determined",
                     "Not known",
                     "Not known if currently on a diet",
                     "Not measured",
                     "Not reported",
                     "Not used",
                     "Not stated")
raw_to_tech <- function(proc, special, types)
{
  var_names <- colnames(proc)
  for (j in 1:length(var_names)) 
  {
    var_val <- var_names[j]
    specials <- special %>%
      filter(variable_name==var_val)
    if (nrow(specials)>0) 
    {
      ind <- which(var_names==var_val)
      var_miss_str  <- paste0(var_val,"_MISS")
      var_miss_reas <- rep("observed",nrow(proc))
      var_vals      <- proc[,ind]
      var_type <- types %>% 
        filter(variable_name==var_val) %>%
        select(variable_type) %>%
        as.character()
      if (var_type=="continuous") {
        for (i in 1:length(var_vals)) {
          if (var_vals[i] %in% specials$value) {
            ind2 <- which(var_vals[i]==specials$value)
            var_vals[i]      <- NA
            var_miss_reas[i] <- specials[ind2,3] %>% as.character()
          }
        }
      }
      if (var_type=="categorical") {
        for (i in 1:length(var_vals)) {
          spec_val  <- specials$value
          spec_meam <- specials$meaning
          if (var_vals[i] %in% spec_val) 
          {
            var_mean <- spec_meam[var_vals[i] == spec_val]
            if (var_mean %in% miss_defs) {
              var_vals[i]      <- NA
              var_miss_reas[j] <- var_mean
            }
          } else {
            var_vals[i]      <- NA
            var_miss_reas[i] <- "unknown"
          }
        }
      }
      if (any(is.na(var_vals))) {
        proc[,ind] <- var_vals
        proc$dummy <- var_miss_reas
        colnames(proc)[ncol(proc)] <- var_miss_str
      }
    }
  }
  return(proc)
}
```

## Raw to technically correct pipeline
```{r}
#read in raw data
raw_biom <- read.csv(here("data","AHS11biomedical.csv"), header=TRUE)
raw_nutr <- read.csv(here("data","AHS11nutrient.csv"), header=TRUE)
raw_food <- read.csv(here("data","AHS11food.csv"), header=TRUE)

quiet_read <- purrr::quietly(readxl::read_excel) # returns [result, output, messages, warning] of a function

# read in dics for raw data
dict_biom <- quiet_read(here("data","nutmstatDataItems2019.xlsx"),sheet=1)$result
dict_nutr <- quiet_read(here("data","nutmstatDataItems2019.xlsx"),sheet=2)$result
dict_food <- quiet_read(here("data","nutmstatDataItems2019.xlsx"),sheet=3)$result

# clean var names
dict_biom <- dict_biom %>% janitor::clean_names() %>% rename(extra=x3)
dict_nutr <- dict_nutr %>% janitor::clean_names() %>% rename(extra1=x3,extra2=x4) 
dict_food <- dict_food %>% janitor::clean_names() %>% rename(extra=x3) 

# remove empty rows
dict_biom <- dict_biom %>% janitor::remove_empty("rows")
dict_nutr <- dict_nutr %>% janitor::remove_empty("rows")
dict_food <- dict_food %>% janitor::remove_empty("rows")

# Create datasets with human readable versions of the variable names
biom_readable <- dict_biom %>% 
  filter(!is.na(variable_name))  
nutr_readable <- dict_nutr %>% 
  filter(!is.na(variable_name)) %>% 
  distinct() # Removes 1 duplicate
food_readable <- dict_food %>% 
  filter(!is.na(variable_name)) %>% 
  distinct()

#Fill in blanks in the variable names
dict_biom <- dict_biom %>% tidyr::fill(variable_name)
dict_nutr <- dict_nutr %>% tidyr::fill(variable_name)
dict_food <- dict_food %>% tidyr::fill(variable_name)  

var_names_biom <- unique(dict_biom$variable_name)
var_names_nutr <- unique(dict_nutr$variable_name)
var_names_food <- unique(dict_food$variable_name)

# Try to infer the data types from the data dictionary
tib1 <- tibble(variable_type=get_type(dict_biom))
tib2 <- tibble(variable_type=rep("continuous",nrow(nutr_readable)))
tib3 <- tibble(variable_type=get_type(dict_food))

# Create a new tibble that takes the readable tibble,
# appends the variable type, and do some minor fixing
types_biom <- bind_cols(biom_readable, tib1) %>%
  mutate(variable_type=ifelse(var_names_biom%in%c("ABSPID","ABSHID"), "string",variable_type))

types_nutr <- bind_cols(nutr_readable, tib2) %>%
  mutate(variable_type=ifelse(variable_name%in%c("ABSPID","ABSHID"), "string",variable_type))

types_food <- bind_cols(food_readable, tib3) %>%
  mutate(variable_type=ifelse(variable_name%in%c("ABSPID","ABSHID"), "string",variable_type)) 
#============================================================
# BIOM DATA
# Made modification to get_special_value_meanings: descriptions for some categorical values (e.g., GLUCFREB) weren't translating correctly (*)
special_biom_new <- get_my_special_value_meanings(dict_biom)
special_biom_new$value <- gsub("\\.", "", special_biom_new$value)  # Remove all dots from the 'value' column

# Remove descriptions
special_biom_new <- special_biom_new %>% 
  filter(!is.na(value)) 

# Convert values to numeric
special_biom_new$value = as.numeric(special_biom_new$value)
#============================================================
#Trying my modified function with the other raw (found it doesnt work):
# special_nutr <- get_my_special_value_meanings(dict_nutr) %>% na.omit()
# special_food <- get_my_special_value_meanings(dict_food) %>% na.omit()
#============================================================
#Using Johns with raw nutr and fodd:
special_nutr <- get_special_value_meanings(dict_nutr) %>% na.omit()
special_food <- get_special_value_meanings(dict_food) %>% na.omit()
#============================================================
# PATCH 3 (*)
miss_defs = tolower(miss_defs) #Adds _MISS vars for COMBCODE and EATOCC
#============================================================

proc_biom <- categorical_to_factor(types_biom,raw_biom)
proc_nutr <- categorical_to_factor(types_nutr,raw_nutr)
proc_food <- categorical_to_factor(types_food,raw_food)

# Convert biom data to technically correct
tech_biom <- raw_to_tech(proc_biom, special_biom_new, types_biom)
tech_nutr <- raw_to_tech(proc_nutr, special_nutr, types_nutr)
tech_food <- raw_to_tech(proc_food, special_food, types_food)
```


PATCH 1 - FIX EXERCISE COLUMNS
```{r}
# Convert 'phys', 'mod', and 'vig' columns to numeric and create corresponding '_MISS' columns
# deliberately keeping 9996 and 9999 and desc in _MISS vars because could need them later

# Convert 'phys', 'mod', and 'vig' columns to numeric and create corresponding '_MISS' columns
tech_biom_mod <- tech_biom %>%
  mutate(
    phys_num = as.numeric(as.character(EXLWTBC)),
    mod_num = as.numeric(as.character(EXLWMBC)),
    vig_num = as.numeric(as.character(EXLWVBC)),
    phys_MISS = ifelse(phys_num %in% c(9996, 9999), as.character(phys_num), "observed"),
    mod_MISS = ifelse(mod_num %in% c(9996, 9999), as.character(mod_num), "observed"),
    vig_MISS = ifelse(vig_num %in% c(9996, 9999), as.character(vig_num), "observed")
  )

# Replace entries in 'phys', 'mod', and 'vig' columns with NA where values are 9996 or 9999
tech_biom_mod$phys_num[tech_biom_mod$phys_num %in% c(9996, 9999)] <- NA
tech_biom_mod$mod_num[tech_biom_mod$mod_num %in% c(9996, 9999)] <- NA
tech_biom_mod$vig_num[tech_biom_mod$vig_num %in% c(9996, 9999)] <- NA


tech_biom = tech_biom_mod %>% select(
  -c(EXLWTBC, EXLWMBC, EXLWVBC)
)
```



PATCH 2 - Link Food Class to Food data
```{r}
df <- tech_food %>%
  select(
    FOODCODC,
    FIVDIG,
    THRDIG,
    TWDIG
  ) %>%
  mutate(
    FOODCODC = as.character(FOODCODC),
    FIVDIG = as.character(FIVDIG),
    THRDIG = as.character(THRDIG),
    TWDIG = as.character(TWDIG)
  )

dict_food_class = quiet_read(here("data","nutmstatDataItems2019.xlsx"),sheet=4)$result


# Shift the column names to the values in the  row
colnames(dict_food_class) <- as.character(dict_food_class[2, ])

# Remove the first row after assigning column names
dict_food_class <- dict_food_class[-c(1,2), ]
dict_food_class <- dict_food_class %>% janitor::clean_names()

# Replace values for FOODCODC
df <- df %>%
  left_join(dict_food_class, by = c("FOODCODC" = "category_code")) %>%
  mutate(FOODCODC = category_label) %>%
  select(-category_label)  # remove the joined 'category_label' column 

# Replace values for FIVDIG
df <- df %>%
  left_join(dict_food_class, by = c("FIVDIG" = "category_code")) %>%
  mutate(FIVDIG = category_label) %>%
  select(-category_label)

# Replace values for THRDIG
df <- df %>%
  left_join(dict_food_class, by = c("THRDIG" = "category_code")) %>%
  mutate(THRDIG = category_label) %>%
  select(-category_label)

# Replace for TWDIG
df <- df %>%
  left_join(dict_food_class, by = c("TWDIG" = "category_code")) %>%
  mutate(TWDIG = category_label) %>%
  select(-category_label)

# Replace original columns in tech_food with the ones from df
tech_food <- tech_food %>%
  mutate(
    FOODCODC = df$FOODCODC,
    FIVDIG = df$FIVDIG,
    THRDIG = df$THRDIG,
    TWDIG = df$TWDIG
  )

```

PATCH 3 - FIXING FOOD DATA
Found that expected _MISS variables are not present in tech_food (corrected by tolower()) (*)
- Other expected _MISS vars (derived from types_food e.g., psaltadd, density, etc.) are not present:
```{r}
# checking if all vars in type_food are present in raw data (found that PSALTADD is not present)
vars_in_types_food = subset(types_food, !is.na(description))$variable_name
vars_in_raw = colnames(raw_food)
setdiff(vars_in_types_food, vars_in_raw)
setdiff(vars_in_raw, vars_in_types_food)
# Can see that the expected vars are not actually present in the raw data +
 #there are 4 redundant cols at the end of raw_food - need to remove these.
tech_food = tech_food %>% select(-setdiff(vars_in_raw, vars_in_types_food))
```

PATCH 4 - MERGING BIOM, NUTR and FOOD

Biom and nutrient can be joined by ABSPID as there are no duplicate entries for PID

```{r}
length(unique(tech_biom$ABSPID))
length(unique(tech_nutr$ABSPID))
```

```{r}
biom_nutr_raw = tech_biom %>%
  full_join(tech_nutr, by = "ABSPID") 
```

FOOD contains multiple entries per person (ABSPID) so its not a trivial join.
Im thinking I create a list object for each ABSPID that contains all the information for that ABSPID.
I.e., Create a column in biom_nutr_raw called FOOD_data ( and for each ABSPID it will contain all info for that ABSPID in tech_food

all info 
```{r}
# Split the tech_food dataframe into a list based on ABSPID
list_data <- with(tech_food, split(tech_food, ABSPID))

# Create a new dataframe with unique ABSPID and corresponding list_data
list_df <- data.frame(ABSPID = names(list_data), FOOD_DATA = I(list_data))

# Merge with biom_nutr_raw
biom_nutr_food <- merge(biom_nutr_raw, list_df, by = "ABSPID", all.x = TRUE)

# If some ABSPIDs from biom_nutr_raw don't have corresponding data in tech_nutr, 
# the FOOD_DATA column for those ABSPIDs will have NAs. If you want to replace NAs with empty lists, you can do:
biom_nutr_food$FOOD_DATA[sapply(biom_nutr_food$FOOD_DATA, is.null)] <- I(list(data.frame()))

# Could further split into daynum for example (there are 2 levels)
```



PATCH 5 - LINK ABORIGINAL AND ISLANDER COHORT

Note that 230ish variables are common between the two datasets. Some variables that are not in the above data dictionary (aboriginal/dictionary.csv) can be found in the numstat dict file (biom, food, nutr, food_class)
```{r}
raw_abor = read.csv(here("data/aboriginal","aboriginal_cohort.csv"), header=TRUE)
dict_abor <- read.csv(here("data/aboriginal","dictionary.csv"),header=TRUE)
```


```{r}
# Function to find descriptions
find_descriptions <- function(col_names, dictionaries) {
  descriptions <- character(0)
  
  for (col_name in col_names) {
    for (dict in dictionaries) {
      match_row <- match(col_name, dict$variable_name)
      if (!is.na(match_row)) {
        descriptions <- c(descriptions, dict$description[match_row])
        break  # Break out of the inner loop once a match is found
      }
    }
  }
  
  return(descriptions)
}
```

Find colnames in common between biom_nutr_food and aboriginal/islander and extract descriptions
```{r}
abor_cols = colnames(raw_abor)
df_cols = colnames(biom_nutr_food)
food_cols = colnames(tech_food)
common_df_abor_cols = intersect(df_cols, abor_cols)
common_food_abor_cols = intersect(food_cols, abor_cols)
#only ABSHID (i.e., food data (in biom_nutr_food) not relev for aboriginal cohort)
descriptions1 <- find_descriptions(common_df_abor_cols, list(types_biom, types_nutr))
descriptions2 <- find_descriptions(common_food_abor_cols, list(types_food))

```


INDGSTAT not in raw data! need another way to classify as indigenous.

<!-- ATTEMPT 1 -->
<!-- the cultural identification sheet  -->
<!-- ```{r} -->
<!-- # set up culture id dictionary for indigenous cohort -->
<!-- dict_abor <- read_excel(here("data/aboriginal", "australian aboriginal and torres strait islander health survey data item list.xls"), sheet = 11, skip = 4) -->
<!-- dict_abor = dict_abor %>% select ( -->
<!--   variable_name = `SAS name`, -->
<!--   `Items & categories`, -->
<!--   `Main population` -->
<!-- )  -->
<!-- dict_abor <- dict_abor %>% janitor::remove_empty("rows") -->
<!-- abor_readable <- dict_abor %>%  -->
<!--   filter(!is.na(variable_name))  -->
<!-- dict_abor <- dict_abor %>% tidyr::fill(variable_name) -->
<!-- var_names_abor <- unique(dict_abor$variable_name) -->
<!-- tib4 <- tibble(variable_type=get_type(dict_abor)) -->
<!-- types_abor <- bind_cols(abor_readable, tib4) -->
<!-- cult_id = types_abor[-c(1,nrow(types_abor)), ] -->
<!-- ``` -->
<!-- Now, can subset aboriginal cohort data based on the cult id related variables (that we can then use to infer indigenous status) -->

<!-- ```{r} -->
<!-- cult_id$variable_name %in% colnames(raw_abor) -->
<!-- 'INDGSTAT' %in% colnames(raw_abor) -->
<!-- ``` -->

Turns out culture vars arent in raw data. Looking at the dictionary.csv file more carefully, i noticed 'Type of Aboriginal and ...Islander household' which led me to the 'household level' sheet in the dic


can use household level data to determine number of aboriginal/islander in each home (only ABSHID is provided in raw data so makes sense)
- rmk: theres a lot of data there e.g., household_type etc that may help with accuracy later on

```{r}
# set up household dictionary for indigenous cohort (might be needed later)
dict_abor <- read_excel(here("data/aboriginal", "australian aboriginal and torres strait islander health survey data item list.xls"), sheet = 4, skip = 4) # skip to where relevant in sheet
dict_abor = dict_abor %>% select (
  variable_name = `SAS name`,
  `Items & categories`,
  `Main population`
) 
dict_abor <- dict_abor %>% janitor::remove_empty("rows")
abor_readable <- dict_abor %>% 
  filter(!is.na(variable_name)) 
dict_abor <- dict_abor %>% tidyr::fill(variable_name)
var_names_abor <- unique(dict_abor$variable_name)
tib4 <- tibble(variable_type=get_type(dict_abor))
types_abor <- bind_cols(abor_readable, tib4)
```


Subset id and number of aboriginal/islander in house
```{r}
df_abor = raw_abor %>% select(
  ABSHID,
  NOATSIHH
)
```


```{r}
# step 1: clean cols that are in biom data
biom_cols = colnames(tech_biom)
common_biom_abor_cols = intersect(biom_cols, abor_cols)
common_biom_abor_cols = c("ABSHID", common_biom_abor_cols)
biom_abor_df = raw_abor[, common_biom_abor_cols]
tech_biom_abor = raw_to_tech(biom_abor_df, special_biom_new, types_biom)
# step 2: clean cols that are in nutr data
nutr_cols = colnames(tech_nutr)
common_nutr_abor_cols = intersect(nutr_cols, abor_cols)
common_nutr_abor_cols = c(common_nutr_abor_cols, "ABSHID")
nutr_abor_df = raw_abor[, common_nutr_abor_cols]
tech_nutr_abor = raw_to_tech(nutr_abor_df, special_nutr, types_nutr)
# step 3: full merge (can filter later if needed)
tech_abor = merge(tech_biom_abor, tech_nutr_abor, by = "ABSHID", all = TRUE)

# Now need to join household_number var
tech_abor = merge(tech_abor, df_abor, by = "ABSHID", all = TRUE)
```


Final DFs
```{r}
head(tech_abor)
head(biom_nutr_food)
```



Save for future use.
```{r}
aborig_household_dict = dict_abor
 save(tech_biom, dict_biom, types_biom,
      tech_nutr, dict_nutr, types_nutr,
      tech_food, dict_food, types_food,
      dict_food_class, biom_nutr_food,
      tech_abor,aborig_household_dict, 
      file = "p5.Rdata")
```




