---
title: "R Notebook"
output: html_notebook
---



Merge biom data for both groups
```{r}
set.seed(123)

biom_ind = types_biom$variable_name[types_biom$variable_name %in% colnames(tech_abor)]
biom_nonind = types_biom$variable_name[types_biom$variable_name %in% colnames(biom_nutr_food)]

names = intersect(biom_ind, biom_nonind)
# subset
ind_biom = tech_abor %>% select(all_of(names)) %>% mutate(group = 'Indigenous')
nonind_biom = biom_nutr_food %>% select(all_of(names)) %>% mutate(group = 'Non-Indigenous')

for (col_name in names(ind_biom)) {
  # Skip the 'group' column as it doesn't need to be changed
  if (col_name != "group") {
    target_class <- class(nonind_biom[[col_name]])
   
    # Check if the classes are different
    if (class(ind_biom[[col_name]]) != target_class) {
      # Coerce to the target class
      if (target_class == "factor") {
        ind_biom[[col_name]] <- as.factor(ind_biom[[col_name]])
      } else if (target_class == "numeric") {
        ind_biom[[col_name]] <- as.numeric(as.character(ind_biom[[col_name]]))  # Convert to character first to avoid potential issues
      } else if (target_class == "integer") {
        ind_biom[[col_name]] <- as.integer(as.character(ind_biom[[col_name]]))  # Convert to character first
      } else if (target_class == "character") {
        ind_biom[[col_name]] <- as.character(ind_biom[[col_name]])
      }  # Add more cases if there are more data types you need to handle
    }
  }
}

# Convert variables in nonind_biom to match those in ind_biom
for (var in names(ind_biom)) {
  if (is.integer(ind_biom[[var]])) {
    nonind_biom[[var]] <- as.integer(as.character(nonind_biom[[var]]))
  }
  else if (is.numeric(ind_biom[[var]])) {
    nonind_biom[[var]] <- as.numeric(as.character(nonind_biom[[var]]))
  }
}

# Combine the dataframes
df = rbind(ind_biom, nonind_biom)
df$group = as.factor(df$group)
df = df %>% filter(AGEC >=18)

# str(df)
```


Transcribe factor levels from dictionary to df
```{r}
for (var in names(df)) {
  # Check the type of the variable from types_biom
  var_type <- subset(types_biom, variable_name == var)$variable_type
 
  if (length(var_type) == 0) {
    next  # skip if the variable is not found in types_biom
  }
 
  if (var_type == "categorical") {
    # Fetch the descriptions for the variable from dict_biom
    descriptions <- subset(dict_biom, variable_name == var)$description
   
    # Extract codes from the descriptions
    codes <- as.numeric(sub("([0-9]+)\\..*", "\\1", descriptions))
   
    # Exclude the rows with NA (which would be the ones that don't have codes)
    valid_rows <- !is.na(codes)
    descriptions <- descriptions[valid_rows]
    codes <- codes[valid_rows]
   
    # The first valid entry after filtering might be a definition, so we should check and remove it
    if(!grepl("^[0-9]+\\.", descriptions[1])){
      descriptions <- descriptions[-1]
      codes <- codes[-1]
    }
   
    # Create a named vector to map the levels
    level_mapping <- setNames(descriptions, codes)
   
    # Convert the variable to a factor and set the levels
    df[[var]] <- factor(df[[var]], levels = as.character(codes), labels = descriptions)
  }
}
# str(df)
```



Create target variables and corresponding criteria that will be used to define CVD risk (labels and corresponding ranges/criteria provided by nutr students)
```{r}

# Filter and create flags
#filtered_df <- df
df <- df %>%
  mutate(
    # Obesity conditions
    Obesity = case_when(
      is.na(BMISC) | is.na(SEX) | is.na(PHDCMWBC) ~ NA_character_,
      BMISC > 30 ~ "Obese",
      (SEX == "1. Male" & PHDCMWBC > 102) | (SEX == "2. Female" & PHDCMWBC > 88) ~ "Obese",
      TRUE ~ "Not Obese"
    ),
   
    # Hypertension conditions
    Hypertension = case_when(
      is.na(SYSTOL) | is.na(DIASTOL) ~ NA_character_,
      SYSTOL > 140 | DIASTOL > 90 ~ "Hypertensive",
      SYSTOL <= 140 & DIASTOL <= 90 ~ "Non-Hypertensive",
    ),
   
    # Diabetes condition
    Diabetes = case_when(
      is.na(DIAHBRSK) ~ NA_character_,
      DIAHBRSK %in% c("2. At risk of diabetes - HbA1c 6.0 - 6.4 %", "3. Indicates diabetes - HbA1c greater than or equal to 6.5%") ~ "Diabetic or at risk",
      DIAHBRSK == "1. Normal - HbA1c <6.0%" ~ "Non Diabetic",
    ),
   
    # ApoB condition
    ApoB = case_when(
      is.na(APOBNTR) ~ NA_character_,
      APOBNTR == "1. Normal ApoB" ~ "Normal ApoB",
      APOBNTR == "2. Abnormal ApoB" ~ "Abnormal ApoB",
    ),
   
    # Dyslipidemia conditions
    Dyslipidemia = case_when(
      # Conditions for Dyslipidemia
      (SEX == "1. Male" & HDLCHREB == "1. Less than 1.0") | (SEX == "2. Female" & HDLCHREB %in% c("1. Less than 1.0", "2. 1.0 to less than 1.3")) ~ "Dyslipidemia",
      !is.na(LDLRESB) & LDLRESB %in% c("06. 3.5 to less than 4.0", "07. 4.0 to less than 4.5", "08. 4.5 or more") ~ "Dyslipidemia",
      !is.na(TRIGRESB) & TRIGRESB %in% c("05. 2.0 to less than 2.5", "2.5 to less than 3.0", "07. 3.0 or more") ~ "Dyslipidemia",
     
      # If no conditions are met and there are no missing variables
      (is.na(SEX) | is.na(HDLCHREB) | is.na(LDLRESB) | is.na(TRIGRESB)) &
      !((SEX == "1. Male" & HDLCHREB == "1. Less than 1.0") | (SEX == "2. Female" & HDLCHREB %in% c("1. Less than 1.0", "2. 1.0 to less than 1.3"))) &
      !(LDLRESB %in% c("06. 3.5 to less than 4.0", "07. 4.0 to less than 4.5", "08. 4.5 or more")) &
      !(TRIGRESB %in% c("05. 2.0 to less than 2.5", "2.5 to less than 3.0", "07. 3.0 or more")) ~ NA_character_,
   
      # Default
      TRUE ~ "No Dyslipidemia"
    ))


   
```



```{r}
cvd_df = df %>% select(ABSPID, SEX, SMKDAILY,
Obesity,
Hypertension,
Diabetes,
ApoB,
Dyslipidemia,
group)
cvd_vars <- c('Obesity', 'Hypertension', 'Diabetes', 'ApoB', 'Dyslipidemia')
str(cvd_df)

```



```{r}
plot_df = cvd_df %>% select(all_of(cvd_vars), group)
gg_miss_fct(plot_df, fct = group)

```




```{r}

levels_of_group <- unique(plot_df$group)

# Loop through each level of the group variable and plot the missingness
for (level in levels_of_group) {
  subset_data <- plot_df[plot_df$group == level, ]
  print(paste("Missingness for group:", level))
  print(gg_miss_upset(subset_data))
}

```


```{r}

filtered_cvd_df <- cvd_df %>%
  filter(
    !(is.na(ApoB) & is.na(Diabetes)), # Not both ApoB and Diabetes are NA
    !(is.na(ApoB) & is.na(Diabetes) & is.na(Dyslipidemia)) # Not all three variables are NA
  )

```

```{r}
plot_df = filtered_cvd_df %>% select(cvd_vars, group)
levels_of_group <- unique(plot_df$group)

# Loop through each level of the group variable and plot the missingness
for (level in levels_of_group) {
  subset_data <- plot_df[plot_df$group == level, ]
  print(paste("Missingness for group:", level))
  print(gg_miss_upset(subset_data))
}

```

```{r}
table(filtered_cvd_df$group)
```

Fix group imbalance (randomly sample 2500 from each)

```{r}

# Split data into Indigenous and Non-Indigenous
indigenous_df <- filter(filtered_cvd_df, group == "Indigenous")
nonindigenous_df <- filter(filtered_cvd_df, group == "Non-Indigenous")

# Randomly select 2500 samples from each group
sampled_indigenous <- sample_n(indigenous_df, 2500)
sampled_nonindigenous <- sample_n(nonindigenous_df, 2500)

# Combine the sampled datasets
sampled_df <- bind_rows(sampled_indigenous, sampled_nonindigenous)
# Convert all columns except ABSPID to factors
sampled_df[, -which(names(sampled_df) == "ABSPID")] <- lapply(sampled_df[, -which(names(sampled_df) == "ABSPID")], factor)

```


compare proportions of 'high cvd risk' in indigenous and Non-Indigenous

```{r}
criteria_proportions <- sampled_df %>%
  group_by(group) %>%
  summarise(
    Total = n(),
    # Current_or_Past_Smoker = sum(SMKDAILY %in% c("1. Current daily smoker", "2. Past daily smoker (still smokes)"), na.rm = TRUE) / Total,
    Obese = sum(Obesity == "Obese", na.rm = TRUE) / Total,
    Hypertensive = sum(Hypertension == "Hypertensive", na.rm = TRUE) / Total,
    Diabetic = sum(Diabetes == "Diabetic or at risk", na.rm = TRUE) / Total,
    Abnormal_ApoB = sum(ApoB == "Abnormal ApoB", na.rm = TRUE) / Total,
    Dyslipidemia = sum(Dyslipidemia == "Dyslipidemia", na.rm = TRUE) / Total
  ) %>%
  select(-Total) %>% rename_all(~gsub("_", " ", .)) #removes '_"



 
# Convert the data to a long format
long_data <- criteria_proportions %>%
  gather(key = "Criteria", value = "Proportion", -group) %>%
  filter(Proportion != 1)  # Exclude proportions that equal 1

# Plot the data
ggplot(long_data, aes(x = Criteria, y = Proportion, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip axes for better visibility with long criterion names
  labs(y = "Proportion",
       x = "Risk factor") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "pink"), name = "Cohort")

```
Appears to be a sig difference in high risk between groups -> (ind w./ higher risk)

lets test this difference:
```{r}
# Initialize an empty vector to store the significant variables
sig_vars <- c()
p_value_df <- data.frame(variable = character(), p_value = numeric())

for (var in cvd_vars) {
  # Subset the data
  x <- sampled_df %>% select(var, group)
  x[[var]] = as.factor(x[[var]])
  # If the variable is a factor, drop the specified levels and perform statistical testing
    # Perform Chi-squared test
    chisq_result <- chisq.test(table(x[[var]], x$group)) 
    p_value_df <- rbind(p_value_df, data.frame(variable = var, p_value = chisq_result$p.value))
   
    # If the p-value is less than 0.05, add the variable name to the sig_vars vector
    # not sure if i should have done a Bonferroni correction? 
    if (chisq_result$p.value < 0.05) {
      sig_vars <- c(sig_vars, var)
    }
}

sorted_biom_p_df <- p_value_df %>% arrange(p_value)

# subset the original dataframe
sig_biom_final <- sampled_df %>% select(ABSPID, group, SEX, SMKDAILY, all_of(sig_vars))
biom_var_info = long_data
```



